Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-03-26 text/paper2/ms.tex
changed exp 4

\begin{abstract}
% Context
The dark-matter halo of the Milky Way is expected to be triaxial on large scales
  and filled with substructure.
It is hoped that streams or shells of stars produced by tidal disruption
  will provide very precise measures of the gravitational potential,
  to test these predictions.
% Aims
We develop a justified method for inferring the gravitational potential
  using precise measurements of stars in tidal streams
  based on an approximate generative model.
Our method does not rely on integrability of the potential;
  we do not know that any good approximation to the large-scale 
  Milky Way potential is integrable.
% Methods: 
Our model is probabilistic, with a likelihood function and priors on the parameters.
It is based on the idea that the stars in a tidal structure
  were close in phase space at some point in the past.
  Model parameters include the present-day phase-space position for each star
  and the progenitor, and thus properly accounts for finite observational uncertainties 
  and missing data.
This method is new in that it makes no assumption about the form of the Milky Way 
  potential, accounts properly for observational uncertainties, and allows for missing data
  dimensions.
We test our method with simulated data in a non-integrable triaxial potential, with realistic 
  observational uncertainties for RR Lyrae stars.
Importantly, the fake data were generated with an n-body simulation,
  not with the approximate model underlying our likelihood function.
% Results:
We find that even with a small number (8 to 16) of well-measured stream stars,
  we can precisely infer properties of the triaxial potential,
  missing phase-space coordinates of the stars,
  and the location of the progenitor object.
The method generalizes to encompass mixtures of independent debris structures,
  or situations in which star membership in structures is unknown.
% [Get more quantitative here?]
% Conclusions: 
\end{abstract}

@@

@@

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the most observable components of a galaxy are sensitive to the inner portion of the host halos mass distribution. For example, the rotation curves of disk galaxies trace the inner mass, especially since matter in disks can be assumed to move on nearly circular orbits. Conversely, measurements of the dark matter distribution at large radii is complicated by the low density of visible tracers, observational difficulties, and assumptions about the smoothness and analytic form. Around external galaxies, the extended mass distribution has been studied using a variety of approaches \citep[see][for a a complete and detailed review]{courteau13}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12c}. 
%\citep{romanowsky03,douglas07,romanowsky09,napolitano11,lee11}
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[see, for example, the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

@@

We identify a number of problems with existing approximate models; each method mentioned above solves only a subset of these issues:
\begin{enumerate}
	\item The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Thus it is critical for any approximate model of tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
	\item The potential of the inner galaxy is conceivably well-described by a (set of) nearly-integrable potentials, thus the dynamics of collisionless systems is plausibly captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. Properly describing the potential may require complex models where computing actions is non-trivial or even impossible.
	\item Near-future photometric surveys such as \gaia\, and the \project{LSST} will likely discover many new streams and kinematic associations of stars. Potential recovery methods should be able to simultaneously model multiple streams and incorporate other dynamical constraints.
	\item Matching generated models to observed stream densities is difficult: models that rely on this must account for observational biases and properly model the underlying background of halo stars. 
	\item Full N-body models are expensive to run. Incorporating an n-body model into a general model and then performing a parameter search would be computationally intensive and, presently, intractable. 
\end{enumerate}

%[][hereafter \paperone]
In \citet{apw13}, we introduced a simple method for using individual stars associated with tidal debris combined with knowledge about the mass and orbit of the progenitor to constrain properties of the host galaxy potential. The method exploits the relationship between the phase-space distribution of debris and (measurable) properties of the progenitor system (e.g., the tidal radius and escape velocity).  Specifically, a better model for the potential is one in which orbits of test particle stars (integrated backwards from their present position) came close (within the tidal radius in position, and escape velocity in velocity) to the orbit of the progenitor. In this article, we use these simple scalings --- which depend only on the mass and orbit of the progenitor and parent potential --- to present a fully probabilistic method (\rewinder) for modeling tidal streams in any underlying potential given data with realistic uncertainties. By ``model'' here we mean a justified, parametrized likelihood function with priors on the parameters. \rewinder\, relies on numerical orbit integration in ordinary phase-space and can thus incorporate arbitrarily complex models for the parent potential. Individual stars are used as constraints, thus \rewinder\, can handle very small samples of well-measured stars (e.g., 4-16). This model trivially extends to incorporating multiple streams, debris structures, and other kinematic information.

@@

\subsection{Tidal debris}\label{sec:debris}

For a satellite galaxy on a mildly eccentric orbit, mass-loss is driven by a combination of tidal stripping due to the steady tidal field of the parent system and tidal shocking due to the rapidly changing tidal field as the progenitor system moves through pericenter \citep[e.g.,][]{choi09}. The interplay between these two processes as a function of orbital properties will be studied in forthcoming work; we focus below on the slow disruption of stars due to steady tidal forcing.

Consider a point-mass satellite of mass $m$ on a circular orbit with frequency $\Omega$ around a more massive mass, $M\gg m$ \citep[the ``restricted three-body problem''; e.g., \S 8.3][]{binneytremaine}. In a frame rotating with the orbital frequency of the satellite, the satellite remains fixed and the static, effective potential around the satellite has (amongst others) two stationary, unstable points located around $\sim R \pm r_J$, where $R$ is the orbital radius of the satellite and $r_J$ is the Jacobi or tidal radius,
\begin{equation}
	r_J \sim R\left(\frac{m}{3M}\right)^{1/3}.\label{eq:ptmass}
\end{equation}
Particles that would be bound to the satellite mass in isolation may have enough (Jacobi) energy to overcome the effective potential barrier at these Lagrange points (Fig.~8.6 of BT) and thus will be preferentially stripped from the satellite at these points. For a spherical, extended parent mass distribution the tidal radius instead scales with the enclosed mass at the instantaneous orbital radius, $R(t)$, with additional terms that account for the local slope of the density profile. The satellite may also be losing mass so that in general the satellite mass also depends on time. The distance scale of the debris then follows
\begin{equation}
	r_{\rm tide}(t) = R(t)\left(\frac{m(t)}{3M_{\rm enc}(R)}\right)^{1/3}\label{eq:tidalradius}
\end{equation}
where $M_{\rm enc}$ is the instantaneous enclosed mass of the parent system within orbital radius $R$.
In general, the effective Lagrange points may not be symmetric about the center of the satellite and may deviate from the tidal radius by factors of order unity.

At pericenter, where a large fraction of the stars are unbound, the velocity of the satellite will be of order $V\sim \sqrt{GM_{\rm enc}/R}$. If the velocity dispersion of the satellite is $\sigma_v \sim \sqrt{Gm/r_J}$, then
\begin{equation}
	\sigma_v \sim V\left(\frac{m}{M_{\rm enc}}\right)^{1/3}\label{eq:velscale}.
\end{equation}
We expect the debris star velocities to also scale with $(m/M_{\rm enc})^{1/3}$. These scalings assume that the satellite is spherical with isotropic velocities; correlations amongst intra-satellite orbits from, e.g., a disk will break these assumptions.

In a triaxial potential, the orbital plane of the satellite is not fixed but we still expect there to be \emph{effective} Lagrange points for a spherical satellite system along the line of centers connecting the origin of the parent potential to the satellite --- that is, we expect the stars to be stripped at some characteristic radius with some dispersion about this point. We proceed by defining a coordinate system relative to the position and velocity of the progenitor, ($\bs{r}_p$,$\bs{v}_p$), rotated into the instantaneous orbital plane defined by $\bs{L} = \bs{r}_p \times \bs{v}_p$. The positions are normalized by the instantaneous tidal radius (Eq.~\ref{eq:tidalradius}) and velocities are normalized by the instantaneous velocity scale (Eq.~\ref{eq:velscale}). 
The (time-dependent) basis vectors are given by
\begin{align}
	\hat{\bs{x}}_1 &= \frac{\bs{r}_p}{\|\bs{r}_p\|}\label{eq:x1}\\
	\hat{\bs{x}}_2 &= \frac{\bs{L} \times \hat{\bs{x}}_1}{\|\bs{L} \times \hat{\bs{x}}_1\|} = \hat{\bs{x}}_3 \times \hat{\bs{x}}_1\\
	\hat{\bs{x}}_3 &= \hat{\bs{L}} = \frac{\bs{L}}{\|\bs{L}\|} = \frac{\bs{r}_p \times \bs{v}_p}{\|\bs{r}_p \times \bs{v}_p\|}\label{eq:x3}.
\end{align}
%(Section~\ref{sec:sims})
Figure~\ref{fig:lpts_r} shows sections of particle orbits from the n-body simulations described above projected into this coordinate system and normalized, symmetric around their unbinding times, $\tub$, for each of the four progenitor masses. In these coordinates, the classical Lagrange points would be located at $x_1\approx\pm1,x_2\approx0,x_3\approx0$ (illustrated by the unit circle in black in Figure~\ref{fig:lpts_r}). Figure~\ref{fig:lpts_v} shows the velocity orbits of each star also projected into these coordinates and normalized. Though not centered on the point-mass Lagrange points, the location of and dispersion about the effective Lagrange points in these scaled coordinates remain remarkably consistent across the range of progenitor masses explored. Furthermore, the dispersion in velocity is well-normalized by the velocity scale of Eq.~\ref{eq:velscale}, though it is clear that the velocity dispersion along $\hat{x}_1$, the radial vector to the satellite position, is larger than that in other dimensions, possibly due to tidal shocking.

We find that even in a triaxial potential, the phase-space dispersion of tidally stripped debris scales with the mass ratio $(m / M_{\rm enc})^{1/3}$. This motivates a model for tidal debris in which each star was ``released'' at the instantaneous, effective Lagrange point at its unbinding time, $\tub$, with some dispersion in position and velocity, all of which depend only on the mass and orbit of the progenitor and the parent potential. We present this model in detail below.

@@

\subsection{Probabilistic model (\rewinder)}
We observe the 6D position of a star, $\D$, in heliocentric coordinates --- e.g., the measured position on the sky, ($l$, $b$); distance, $d$; proper motions, ($\mu_l$, $\mu_b$); and line-of-sight velocity, $v_r$ --- and have determined through some other means that this star was once part of a progenitor system (e.g., satellite galaxy) with mass $m(t)$ that is disrupting and forming a cold debris structure in the potential, $\Phi$, of the parent galaxy (e.g., the Milky Way). We assume that the mass of the parent system enclosed within the pericenter of the orbit of the satellite, $M_{\rm enc}(R_{\rm peri})$, is much larger than the initial mass of the satellite, $M_{\rm enc}(R_{\rm peri})\gg m(t=0)$. The present position of the progenitor is observed to be at heliocentric position $\D_\sat$ (where any subscript $\sat$ refers to the progenitor). In general, the data for the star and progenitor will have missing dimensions and significant uncertainties at distances typical to the Galactic halo, thus we define $\W$ and $\W_\sat$ as the true 6D, heliocentric positions of the star and progenitor and will include these in our model. By including these positions in the model, this model will work even when the star has missing dimensions or the progenitor location is unknown as in, e.g., the Orphan stream \citep{belokurov07}.

@@

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r},\bs{v}$) and ($\bs{r}_\sat,\bs{v}_\sat$) are the position and velocity of the star and progenitor. In cartesian coordinates, the present-day position of the star is taken as having been sampled at $t=\tub$ from an isotropic Gaussian centered on one of the instantaneous, effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity, evolved to present day in the parent potential, $\Phi$, parametrized by the parameter vector $\bs{\Theta}_\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored, and thus treat the star as a test particle.\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~$M$_\odot$.} The satellite mass enters the model through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m_0$, and a constant mass-loss term, $\dot{m}$ into the model: $m(t) = m_0 - \dot{m}t$. Finally, we add two additional parameters: a constant, global offset, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. To compress notation, we pack all progenitor parameters ($\W_\sat, m_0, \dot{m}, \Loffset$) into the vector $\bs{\Theta}_\sat$, and all star parameters ($\W, \tailbit, \tub$) into the vector $\bs{\Theta}$. The likelihood is then:

@@

For all experiments below, we use a Markov Chain Monte Carlo (MCMC) algorithm to sample from the posterior probability distribution given by our model. Standard MCMC algorithms (e.g., Metropolis-Hastings) update a single chain while exploring parameter space. We instead use an affine-invariant ``ensemble'' sampler \citep{goodman10} that, each step in parameter space, updates the positions of many ``walkers'' (the ensemble). This algorithm is implemented in the \project{Python} programming language \citep{foremanmackey13} and runs naturally in a parallelized environment, e.g., the message passing interface (MPI). In each experiment we run the walkers for a large number of steps from starting positions described below, then throw away these samples and take the positions from the final step of this ``burn-in'' phase as a starting point for the samples used for inference. We compute the autocorrelation time for each sampled parameter (using \texttt{ACOR}\footnote{www.math.nyu.edu/faculty/goodman/software/acor/}$^{,}$\footnote{https://github.com/dfm/acor}) and thin the chains by taking every $\mathrm{median}(t_{\rm acor})$ sample to ensure the samples are close to independent.

% TODO: m and mdot!!
\begin{table}[h]
\begin{center}
	\begin{tabular}{l c l} \toprule
		\multicolumn{3}{l}{{\bf \emph{Milky Way parameters} ($\bs{\Theta}_\Phi$)}} \\
		\toprule
		Component & Parameter & Prior \\\toprule
		disk & $M_{\rm disk}$ & $1.0\times10^{11}M_\odot$ (fixed) \\ 
		& $a$ & 6.5 kpc (fixed)\\
		& $b$ & 0.26 kpc (fixed)\\
		\midrule
		spheroid & $M_{\rm spher}$ & $3.4\times10^{10}M_\odot$ (fixed)\\ 
		& $c$ & 0.7 kpc (fixed)\\
		\midrule
		halo & $\vhalo$ & $\mathcal{U}(100,200)$ km/s \\
		& $q_1$ & $\mathcal{U}(1,2)$\\
		& $q_2$ & 1.0 (fixed)\\
		& $q_z$ & $\mathcal{U}(1,2)$\\
		& $\phi$ & $\mathcal{U}(52,142)$ deg\\
		& $r_h$ & 12 kpc (fixed)\\
		\toprule
		\multicolumn{3}{l}{{\bf \emph{Progenitor parameters} ($\bs{\Theta}_\sat$)}} \\
		\toprule
		position & $\bs{r}_\sat$ & $\|\bs{r}_\sat\|\sim\mathcal{U}(0,200)$~kpc \\
		velocity & $\bs{v}_\sat$ & $\|\bs{v}_\sat\|\sim\mathcal{N}(\mu=0,\sigma^2=4000)$~km/s\\
		Lagrange pt. offset & $\alpha$ & $\mathcal{N}(\mu=2,\sigma^2=1)$\\
		\toprule
		\multicolumn{3}{l}{{\bf \emph{Star parameters} ($\bs{\Theta}^{(i)}$)}} \\
		\toprule
		position & $\bs{r}$ & $\|\bs{r}\|\sim\mathcal{U}(0,200)$~kpc \\
		velocity & $\bs{v}$ & $\|\bs{v}\|\sim\mathcal{N}(\mu=0,\sigma^2=4000)$~km/s\\
		tail assignment (binary) & $\beta$ & $\pm1$~equally likely\\ % TODO: how to say equal weight to -1 or 1?
%		prob. shocked & $\pshock$ & $\mathcal{U}(0,1)$\\
		time unbound & $\tub$ & $\mathcal{U}(0,\tint)$\\ % TODO: talk about \tint (integration time) in text!
		\bottomrule
		\end{tabular}
	\caption{Parameter values used ...\label{tbl:params}}
\end{center}
\end{table}
% TODO: in caption, redefine N is normal U is uniform. Star parameters are repeated for each star. summarize number of parameters for milky way, progenitor, etc. -- how to read the parameter

@@

>>>>

\begin{abstract}
% Context
The dark-matter halo of the Milky Way is expected to be triaxial on large scales
  and filled with substructure.
It is hoped that streams or shells of stars produced by tidal disruption
  will provide very precise measures of the gravitational potential,
  to test these predictions.
% Aims
We develop a justified method for inferring the gravitational potential
  using precise measurements of stars in tidal streams
  based on an approximate generative model.
Our method does not rely on integrability of the potential;
  we do not know that any good approximation to the large-scale 
  Milky Way potential is integrable.
% Methods: 
Our model is probabilistic, with a likelihood function and priors on the parameters.
It is based on the idea that the stars in a tidal structure
  were close in phase space at some point in the past.
  Model parameters include the present-day phase-space position for each star
  and the progenitor, and thus properly accounts for finite observational uncertainties 
  and missing data.
This method is new in that it makes no assumption about the form of the Milky Way 
  potential, accounts properly for observational uncertainties, and allows for missing data
  dimensions.
We test our method with simulated data in a non-integrable triaxial potential, with realistic 
  observational uncertainties for RR Lyrae stars.
Importantly, the fake data were generated with an n-body simulation,
  not with the approximate model underlying our likelihood function.
% Results:
We find that even with a small number (8) of well-measured stream stars,
  we can precisely infer properties of the triaxial potential and
  missing phase-space coordinates of the stars.
The method generalizes to encompass mixtures of independent debris structures,
  or situations in which star membership in structures is unknown.
% [Get more quantitative here?]
% Conclusions: 
\end{abstract}

@@

% TODO: audit all uses of the word 'model'

@@

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the most observable components of a galaxy are sensitive to the inner portion of the host halos mass distribution. For example, the rotation curves of disk galaxies trace the inner mass, especially since matter in disks can be assumed to move on nearly circular orbits. Conversely, measurements of the dark matter distribution at large radii is complicated by the low density of visible tracers, observational difficulties, and assumptions about the smoothness and analytic form. Around external galaxies, the extended mass distribution has been studied using a variety of approaches \citep[see][for a a complete and detailed review]{courteau13}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12c}. 
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[see, for example, the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

@@

We identify a number of problems with existing approximate models; each method mentioned above solves only a subset of these issues:
\begin{enumerate}
	\item The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Thus it is critical for any approximate model of tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
	\item The potential of the inner galaxy is conceivably well-described by a (set of) nearly-integrable potentials, thus the dynamics of collisionless systems is plausibly captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. Properly describing the potential may require complex models where computing actions is non-trivial or even impossible.
	\item Near-future photometric surveys such as \gaia\, and the \project{LSST} will likely discover many new streams and kinematic associations of stars. Potential recovery methods should be able to simultaneously model multiple streams and incorporate other dynamical constraints.
	\item Matching generated models to observed stream densities is difficult: models that rely on this must account for observational biases and properly model the underlying background of halo stars. 
	\item Full N-body simulations are expensive to run. Incorporating an n-body model into a general model and then performing a parameter search would be computationally intensive and, presently, intractable. 
\end{enumerate}

In \citet{apw13}, we introduced a simple method for using individual stars associated with tidal debris combined with knowledge about the mass and orbit of the progenitor to constrain properties of the host galaxy potential. The method exploits the relationship between the phase-space distribution of debris and (measurable) properties of the progenitor system (e.g., the tidal radius and escape velocity).  Specifically, a better model for the potential is one in which orbits of test particle stars (integrated backwards from their present position) came close (within the tidal radius in position, and escape velocity in velocity) to the orbit of the progenitor. In this article, we use these simple scalings --- which depend only on the mass and orbit of the progenitor and parent potential --- to present a fully probabilistic method (\rewinder) for modeling tidal streams in any underlying potential given data with realistic uncertainties. By ``model'' we mean a justified, parametrized likelihood function with priors on the parameters. \rewinder\, relies on numerical orbit integration in ordinary phase-space and can thus incorporate arbitrarily complex models for the parent potential. Individual stars are used as constraints, thus \rewinder\, can handle very small samples of well-measured stars (e.g., 4-16). This model trivially extends to incorporating multiple streams, debris structures, and other kinematic information.

@@

\subsection{Motivation: tidal debris}\label{sec:debris}

For a satellite galaxy on a mildly eccentric orbit, mass loss is driven by a combination of tidal stripping by the steady tidal field of the parent system and tidal shocking by the rapidly changing tidal field as the progenitor system moves through pericenter \citep[e.g.,][]{choi09}. The interplay between these two processes as a function of orbital properties is outside our scope; we focus here on the slow removal of stars by steady tidal forcing.

Consider a point-mass satellite of mass $m$ on a circular orbit with frequency $\Omega$ around a more massive mass $M\gg m$ \citep[the ``restricted three-body problem''; e.g., \S 8.3][]{binneytremaine}. In a frame rotating with the orbital frequency of the satellite, the satellite remains fixed and the static, effective potential around the satellite has (amongst others) two stationary, unstable points located around galactocentric radii $\sim R \pm r_J$, where $R$ is the orbital radius of the satellite and $r_J$ is the Jacobi or tidal radius
\begin{equation}
	r_J \sim R\left(\frac{m}{3M}\right)^{1/3}.\label{eq:ptmass}
\end{equation}
Particles that would be bound to the satellite in isolation may have enough (Jacobi) energy to overcome the effective potential barrier at these Lagrange points (Fig.~8.6 of BT) and thus will be preferentially stripped from the satellite at these points. For a spherical, extended parent mass distribution the tidal radius instead scales with the enclosed mass at the instantaneous orbital radius, $R(t)$, with additional terms that account for the local slope of the density profile. In general the satellite mass also depends on time. The distance scale of the debris then follows
\begin{equation}
	r_{\rm tide}(t) = R(t)\left(\frac{m(t)}{3M_{\rm enc}(R)}\right)^{1/3}\label{eq:tidalradius}
\end{equation}
where $M_{\rm enc}$ is the instantaneous enclosed mass of the parent system within orbital radius $R$.
In general, the effective Lagrange points may not be symmetric about the center of the satellite and may deviate from the tidal radius by factors of order unity.

At pericenter, where a large fraction of the stars are unbound, the velocity of the satellite will be of order $V\sim \sqrt{GM_{\rm enc}/R}$. If the velocity dispersion of the satellite is $\sigma_v \sim \sqrt{Gm/r_J}$, then
\begin{equation}
	\sigma_v \sim V\left(\frac{m}{M_{\rm enc}}\right)^{1/3}\label{eq:velscale}.
\end{equation}
We expect the debris-star velocities also to scale with $(m/M_{\rm enc})^{1/3}$. These scalings assume that the satellite is spherical with isotropic velocities; correlations among intra-satellite orbits from, e.g., a disk will break these assumptions.

In a triaxial potential, the orbital plane of the satellite is not fixed but we still expect there to be \emph{effective} Lagrange points for a spherical satellite system along the line of centers connecting the origin of the parent potential to the satellite --- that is, we expect the stars to be stripped at some characteristic radius with some dispersion about this point. We proceed by defining a coordinate system relative to the position and velocity of the progenitor, ($\bs{r}_p$,$\bs{v}_p$), rotated into the instantaneous orbital plane defined by $\bs{L} = \bs{r}_p \times \bs{v}_p$. The positions are normalized by the instantaneous tidal radius (Eq.~\ref{eq:tidalradius}) and velocities are normalized by the instantaneous velocity scale (Eq.~\ref{eq:velscale}). 
The (time-dependent) basis vectors are given by
\begin{align}
	\hat{\bs{x}}_1 &= \frac{\bs{r}_p}{\|\bs{r}_p\|}\label{eq:x1}\\
	\hat{\bs{x}}_2 &= \frac{\bs{L} \times \hat{\bs{x}}_1}{\|\bs{L} \times \hat{\bs{x}}_1\|} = \hat{\bs{x}}_3 \times \hat{\bs{x}}_1\\
	\hat{\bs{x}}_3 &= \hat{\bs{L}} = \frac{\bs{L}}{\|\bs{L}\|} = \frac{\bs{r}_p \times \bs{v}_p}{\|\bs{r}_p \times \bs{v}_p\|}\label{eq:x3}.
\end{align}
Figure~\ref{fig:lpts_r} shows sections of particle orbits (for particles that are stripped) from the n-body simulations described above projected into this coordinate system and normalized, symmetric around their unbinding times, $\tub$, for each of the four progenitor masses. In these coordinates, the classical Lagrange points would be located at $x_1\approx\pm\rtide,x_2\approx0,x_3\approx0$ (illustrated by the intersection of the black unit circles and horizontal lines in Figure~\ref{fig:lpts_r}). Figure~\ref{fig:lpts_v} shows the velocity orbits of each star also projected into these coordinates and normalized. Though not centered on the point-mass Lagrange points, the location of and dispersion about the effective Lagrange points in these scaled coordinates remain remarkably consistent across the range of progenitor masses explored. Furthermore, the dispersion in velocity is well-normalized by the velocity scale of Eq.~\ref{eq:velscale}, though it is clear that the velocity dispersion along $\hat{x}_1$, the radial vector to the satellite position, is larger than that in other dimensions, possibly because of tidal shocking.

We find that even in a triaxial potential, the dispersion --- in position and velocity --- of tidally stripped debris as it comes unbound from the satellite scales with the mass ratio $(m / M_{\rm enc})^{1/3}$. This motivates a model for tidal debris in which each star was ``released'' at the instantaneous, effective Lagrange point at its unbinding time, $\tub$, with a dispersion in position and velocity, all of which depend only on the mass and orbit of the progenitor and the parent potential. We present this model in detail below.

@@

\subsection{Probabilistic model (\rewinder)}
We observe the 6D position of a star, $\D$, in heliocentric coordinates --- e.g., the measured position on the sky, ($l$, $b$); distance, $d$; proper motions, ($\mu_l$, $\mu_b$); and line-of-sight velocity, $v_r$ --- and have determined through some other means that this star was once part of a progenitor system (e.g., satellite galaxy) with mass $m(t)$ that is disrupting and forming a cold debris structure in the potential, $\Phi$, of the parent galaxy (e.g., the Milky Way). We assume that the mass of the parent system enclosed within the pericenter of the orbit of the satellite, $M_{\rm enc}(R_{\rm peri})$, is much larger than the initial mass of the satellite, $M_{\rm enc}(R_{\rm peri})\gg m(t=0)$. The present position of the progenitor is observed to be at heliocentric position $\D_\sat$ (where any subscript $\sat$ refers to the progenitor). In general, the data for the star and progenitor will have missing dimensions and significant uncertainties at distances typical to the Galactic halo, thus we define $\W$ and $\W_\sat$ as the true 6D, heliocentric positions of the star and progenitor and will include these in our model. Since we include these positions as parameters, this model will work even when the star has missing dimensions or the progenitor location is unknown as in, e.g., the Orphan stream \citep{belokurov07}.

@@

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r},\bs{v}$) and ($\bs{r}_\sat,\bs{v}_\sat$) are the position and velocity of the star and progenitor. In cartesian coordinates, the present-day position of the star is taken as having been sampled at $t=\tub$ from an isotropic Gaussian centered on one of the instantaneous, effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity, evolved to present day in the parent potential, $\Phi$, parametrized by the parameter vector $\bs{\Theta}_\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored,\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~$M$_\odot$.} and thus treat the star as a test particle. The satellite mass enters the model through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m_0$, and a constant mass-loss term, $\dot{m}$ into the model: $m(t) = m_0 - \dot{m}t$. Finally, we add two additional parameters: a constant, global offset, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. To compress notation, we pack all progenitor parameters ($\W_\sat, m_0, \dot{m}, \Loffset$) into the vector $\bs{\Theta}_\sat$, and all star parameters ($\W, \tailbit, \tub$) into the vector $\bs{\Theta}$. The likelihood is then:

@@

For all experiments below, we use a Markov Chain Monte Carlo (MCMC) algorithm to sample from the posterior probability distribution given by our model. Standard MCMC algorithms (e.g., Metropolis-Hastings) update a single chain while exploring parameter space. We instead use an affine-invariant ``ensemble'' sampler \citep{goodman10} that, each step in parameter space, updates the positions of many ``walkers'' (the ensemble). This algorithm is implemented in the \project{Python} programming language \citep{foremanmackey13} and runs naturally in a parallelized environment such as the message passing interface (MPI). In each experiment we run the walkers for a large number of steps from starting positions described below, then throw away these samples and take the positions from the final step of this ``burn-in'' phase as a starting point for the samples used for inference. We compute the autocorrelation time for each sampled parameter (using \texttt{ACOR}\footnote{www.math.nyu.edu/faculty/goodman/software/acor/}$^{,}$\footnote{https://github.com/dfm/acor}) and thin the chains by taking every $\mathrm{median}(t_{\rm acor})$ sample to ensure the samples are close to independent.

\begin{table}[h]
\begin{center}
	\begin{tabular}{l c l} \toprule
		\multicolumn{3}{l}{{\bf \emph{Milky Way parameters} ($\bs{\Theta}_\Phi$)}} \\
		\toprule
		Component & Parameter & Prior \\\toprule
		disk & $M_{\rm disk}$ & $1.0\times10^{11}M_\odot$ (fixed) \\ 
		& $a$ & 6.5 kpc (fixed)\\
		& $b$ & 0.26 kpc (fixed)\\
		\midrule
		spheroid & $M_{\rm spher}$ & $3.4\times10^{10}M_\odot$ (fixed)\\ 
		& $c$ & 0.7 kpc (fixed)\\
		\midrule
		halo & $\vhalo$ & $\mathcal{U}(100,200)$ km/s \\
		& $q_1$ & $\mathcal{U}(1,2)$\\
		& $q_2$ & 1.0 (fixed)\\
		& $q_z$ & $\mathcal{U}(1,2)$\\
		& $\phi$ & $\mathcal{U}(52,142)$ deg\\
		& $r_h$ & 12 kpc (fixed)\\
		\toprule
		\multicolumn{3}{l}{{\bf \emph{Progenitor parameters} ($\bs{\Theta}_\sat$)}} \\
		\toprule
		position & $\bs{r}_\sat$ & $\|\bs{r}_\sat\|\sim\mathcal{U}(0,200)$~kpc \\
		velocity & $\bs{v}_\sat$ & $\|\bs{v}_\sat\|\sim\mathcal{N}(\mu=0,\sigma^2=4000)$~km/s\\
		Lagrange pt. offset & $\alpha$ & $\mathcal{N}(\mu=2,\sigma^2=1)$\\
		initial mass & $m_0$ & $2.5\times10^8~M_\odot$ (fixed)\\
		mass loss & $\dot{m}$ & $3.2\times10^4~M_\odot$/Myr (fixed)\\
		\toprule
		\multicolumn{3}{l}{{\bf \emph{Star parameters} ($\bs{\Theta}^{(i)}$)}} \\
		\toprule
		position & $\bs{r}$ & $\|\bs{r}\|\sim\mathcal{U}(0,200)$~kpc \\
		velocity & $\bs{v}$ & $\|\bs{v}\|\sim\mathcal{N}(\mu=0,\sigma^2=4000)$~km/s\\
		tail assignment (binary) & $\beta$ & $\pm1$~equally likely (fixed at truth)\\ % TODO: how to say equal weight to -1 or 1?
%		prob. shocked & $\pshock$ & $\mathcal{U}(0,1)$\\
		time unbound & $\tub$ & $\mathcal{U}(0,6.2)~Gyr$\\ % TODO: talk about \tint (integration time) in text!
		\bottomrule
		\end{tabular}
	\caption{Parameter values used in the experiments of Section~\ref{sec:experiments}. $\mathcal{N}$ is the normal (Gaussian) distribution, and $\mathcal{U}$ the uniform distribution. There are 11 parameters for the Milky Way potential, but only four are left free to vary; some parameters are fixed (denoted by ``(fixed)'') at the true values used in the n-body simulations that generated the fake data. The progenitor has nine parameters --- the position, $\bs{r}_{\rm p}$, and velocity, $\bs{v}_{\rm p}$, vectors each contain three components --- but only five are left free to vary. The sky coordinates (e.g., Galactic $l$, $b$) are assumed to be known with negligible uncertainty. Each star has eight associated parameters, five of which are allowed to vary. Sky coordinates are fixed, along with the tail assignment (whether the star belongs to the leading or trailing tail). For inference with eight stars, there are $4+5+8\times5=49$ free parameters. \label{tbl:params}}
\end{center}
\end{table}

@@

% TODO: slight ~1% bias in values due to particular sample of stars?
